---
title: "PROJECT: HEART DISEASE PREDICTION USING MACHINE LEARNING"
output: html_document
date: "2025-04-29"
---

# Heart Disease Prediction using Machine Learning

**Import all relevant libraries**

```{r}
library(dplyr)
library(caret)
library(ggplot2)
library(ggcorrplot)
library(glmnet)
library(mltools)
library(data.table)
library(keras)
library(tfruns)
library(tidyr)
library(GGally)
library(corrplot)
library(vimp)
library(graphics)
library(pROC)
```

**1. Explore the overall structure of the data**

First read the data

```{r}
heart_data = read.csv('heart_disease_uci.csv')
head(heart_data)
```

**1.1 Check the structure of the data**

```{r}
str(heart_data)
```

**1.2 Summary of the data**

```{r}
# Summary of the data
summary(heart_data)
```

The dataset includes variables such as:

> `age`: Age in years
>
> `sex`: (1 = male, 0 = female)
>
> `cp`: Chest pain type (4 categories)
>
> `trestbps`: Resting blood pressure (mm Hg)
>
> `chol`: Serum cholesterol (mg/dl)
>
> `fbs`: Fasting blood sugar \> 120 mg/dl (1 = true, 0 = false)
>
> `restecg`: Resting electrocardiographic results (3 categories)
>
> `thalach`: Maximum heart rate achieved
>
> `exang`: Exercise-induced angina (1 = yes, 0 = no)
>
> `oldpeak`: ST depression induced by exercise
>
> `slope`: Slope of the peak exercise ST segment (3 categories)
>
> `ca`: Number of major vessels (0-3) colored by fluoroscopy
>
> `thal`: Thalassemia condition (3 categories)
>
> `num`: Target (0 = no heart disease, 1, 2, 3, 4 = heart disease)

```{r}
dim(heart_data)
```

**1.3 Check for missing values**

```{r}
# Calculate the number of missing values per column
null_counts <- colSums(is.na(heart_data))

# Create a data frame for plotting
null_df <- data.frame(
  Column = names(null_counts),
  Missing_Values = as.numeric(null_counts)
)

# Filter out columns with zero missing values (optional, for cleaner plot)
null_df <- null_df[null_df$Missing_Values > 0, ]

null_df
```

```{r}
# Create a bar plot of missing values
ggplot(data = null_df, aes(x = reorder(Column, -Missing_Values), y = Missing_Values, fill = Column)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  labs(
    title = "Missing Values by Column in Heart Disease Dataset",
    x = "Column",
    y = "Number of Missing Values"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

**2. Exploratory Data Analysis**

**2.1 Target Distribution**

The target variable has 5 classes:

-   **0**: No heart disease

-   **1**: Mild heart disease

-   **2**: Moderate heart disease

-   **3**: Severe heart disease

-   **4**: Very severe heart disease

We will check how many patients belong to each class to understand the overall class distribution in the dataset.

```{r}
# Distribution of the target variable (num, 5 classes)
target_counts <- table(heart_data$num)
cat("Target Variable Distribution:\n")
print(target_counts)

```

```{r}
ggplot(data = heart_data, aes(x = as.factor(num), fill = as.factor(num))) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(
    title = "Distribution of Heart Disease Severity (Target Variable)",
    x = "Severity Classes (0 = No Heart Disease, 1-4 = Severity Levels)",
    y = "Count"
  ) +
  theme_minimal()
```

**2.2 Number of Patients in Each Class**

Check number of patients in each class.

```{r}
# Count of patients with and without heart disease
num_with_disease <- sum(heart_data$num %in% c(1, 2, 3, 4))
num_without_disease <- sum(heart_data$num == 0)

# Display the count of patients with and without heart disease
cat("### Count of Patients with and without Heart Disease ###\n")
cat("Number of patients with heart disease:", num_with_disease, "\n")
cat("Number of patients without heart disease:", num_without_disease, "\n\n")

# Total number of patients in the dataset
total <- nrow(heart_data)

# Distribution of the target variable
target_counts <- table(heart_data$num)

# Display the distribution of the target variable with percentages
cat("Target Variable Distribution (Heart Disease Severity Classes) \n")
for (class_label in names(target_counts)) {
  count <- target_counts[class_label]
  percentage <- (count / total) * 100
  cat(sprintf("Class %s: %d patients (%.2f%%)\n", class_label, count, percentage))
}
```

This shows that the data is imbalanced. To address this we can follow two approaches

1.  First one is to use the SMOTE and it will help us increase the minority samples

2.  In second one we can combine all of the abormal types to a single category and make this problem a binary classification.

**3. Demographic Insights (`Age` and `Gender`)**

**3.1 Age**

```{r}
min_age <- min(heart_data$age, na.rm = TRUE)
max_age <- max(heart_data$age, na.rm = TRUE)
age_range <- max_age - min_age

# Display the age range
cat(" Age Range of Patients\n")
cat("Minimum age:", min_age, "\n")
cat("Maximum age:", max_age, "\n")
cat("Age range:", age_range, "years\n\n")
```

```{r}
# Mean, median, and mode of age column
age_mean <- mean(heart_data$age, na.rm = TRUE)
age_median <- median(heart_data$age, na.rm = TRUE)
age_mode <- as.numeric(names(sort(table(heart_data$age), decreasing = TRUE)[1]))

# Display mean, median, and mode
cat("Mean:", age_mean, "\n")
cat("Median:", age_median, "\n")
cat("Mode:", age_mode, "\n\n")

# Histogram of age with mean, median, and mode lines
library(ggplot2)
ggplot(data = heart_data, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  geom_density(alpha = 0.2, fill = "blue") +
  geom_vline(aes(xintercept = age_mean), color = "red", linetype = "dashed") +
  geom_vline(aes(xintercept = age_median), color = "green", linetype = "dashed") +
  geom_vline(aes(xintercept = age_mode), color = "blue", linetype = "dashed") +
  labs(
    title = "Age Distribution of Patients",
    x = "Age",
    y = "Density"
  ) +
  theme_minimal()
```

The 49-year age range (28 to 77 years) reflects a diverse patient population, critical for understanding heart disease risk across age groups.

This broad span suggests models must account for varying age-related risk factors.

**3.2 Gender**

```{r}
# Assuming sex is coded as "Male"/"Female" or "M"/"F"; adjust if coded as 1/0
gender_counts <- table(heart_data$sex)

# Display gender distribution
cat("### Gender Distribution of the Patients ###\n")
cat("Male patients:", gender_counts[["Male"]] %||% gender_counts[["M"]], "\n")
cat("Female patients:", gender_counts[["Female"]] %||% gender_counts[["F"]], "\n\n")

```

```{r}
# Bar plot for gender distribution
ggplot(data = heart_data, aes(x = sex, fill = sex)) +
  geom_bar() +
  scale_fill_manual(values = c("Male" = "#66C2A5", "Female" = "#FC8D62")) +
  labs(
    title = "Gender Distribution of Patients",
    x = "Gender",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# Histogram of age by gender
ggplot(data = heart_data, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge", bins = 30) +
  scale_fill_manual(values = c("Male" = "#66C2A5", "Female" = "#FC8D62")) +
  labs(
    title = "Age Distribution by Gender",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()
```

The predominance of male patients (726 vs. 194 females) highlights a gender imbalance that may influence model performance.

Further analysis is needed to assess if heart disease patterns differ by gender in this dataset.

**4. Health Metrics**

```{r}
# Calculate average and median values for resting blood pressure (trestbps)
avg_trestbps <- mean(heart_data$trestbps, na.rm = TRUE)
median_trestbps <- median(heart_data$trestbps, na.rm = TRUE)

# Calculate average and median values for serum cholesterol (chol)
avg_chol <- mean(heart_data$chol, na.rm = TRUE)
median_chol <- median(heart_data$chol, na.rm = TRUE)

# Calculate average and median values for maximum heart rate achieved (thalch)
avg_thalach <- mean(heart_data$thalch, na.rm = TRUE)
median_thalach <- median(heart_data$thalch, na.rm = TRUE)

# Display the results
cat("Health Metrics\n")
cat(sprintf("Average Resting Blood Pressure (trestbps): %.2f\n", avg_trestbps))
cat(sprintf("Median Resting Blood Pressure (trestbps): %.0f\n", median_trestbps))
cat(sprintf("Average Serum Cholesterol (chol): %.2f\n", avg_chol))
cat(sprintf("Median Serum Cholesterol (chol): %.0f\n", median_chol))
cat(sprintf("Average Maximum Heart Rate Achieved (thalch): %.2f\n", avg_thalach))
cat(sprintf("Median Maximum Heart Rate Achieved (thalch): %.0f\n\n", median_thalach))

```

```{r}
# Combine relevant columns for plotting
health_metrics <- heart_data[, c("trestbps", "chol", "thalch")]

# Reshape data to long format for boxplot
health_metrics_long <- pivot_longer(health_metrics, cols = everything(), names_to = "Metric", values_to = "Value")

# Create the boxplot
ggplot(data = health_metrics_long, aes(x = Metric, y = Value, fill = Metric)) +
  geom_boxplot() +
  scale_fill_manual(values = c("trestbps" = "#66C2A5", "chol" = "#FC8D62", "thalch" = "#8DA0CB")) +
  labs(
    title = "Comparison of Health Metrics: Resting BP, Serum Cholesterol, and Max Heart Rate",
    x = "Health Metrics",
    y = "Values"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

**Resting Blood Pressure** **(trestbps)**: The average (132.13 mmHg) and median (130 mmHg) suggest a typical range, but numerous outliers indicate potential hypertension cases requiring closer examination.

**Serum** **Cholesterol (chol)**: A lower average (199.13 mg/dl) than median (223 mg/dl) with many outliers points to a skewed distribution, highlighting patients with extreme cholesterol levels.

**Maximum** **Heart Rate (thalch)**: The consistent average (137.55 bpm) and median (140 bpm) with fewer outliers suggest stable exercise responses across patients.

**5. Categorical Features**

**5.1 Exercise-Induced Angina (exang)**

```{r}
#Count of patients with exercise-induced angina (TRUE/FALSE)
exang_counts <- table(heart_data$exang, useNA = "ifany")
exang_percentage <- prop.table(exang_counts) * 100

# Display counts and percentages for exercise-induced angina
cat("Exercise-Induced Angina (exang)\n")
cat(sprintf("Patients with exercise-induced angina (TRUE): %d (%.2f%%)\n", 
            exang_counts["TRUE"], exang_percentage["TRUE"]))
cat(sprintf("Patients without exercise-induced angina (FALSE): %d (%.2f%%)\n\n", 
            exang_counts["FALSE"], exang_percentage["FALSE"]))
```

```{r}
# Plot for Exercise-Induced Angina (exang)
ggplot(data = heart_data, aes(x = exang, fill = exang)) +
  geom_bar() +
  scale_fill_manual(values = c("FALSE" = "#66C2A5", "TRUE" = "#FC8D62")) +
  labs(
    title = "Distribution of Exercise-Induced Angina",
    x = "Exercise-Induced Angina (exang)",
    y = "Number of Patients"
  ) +
  scale_x_discrete(labels = c("FALSE" = "No", "TRUE" = "Yes")) +
  theme_minimal() +
  theme(legend.position = "none")
```

Exercise-Induced Angina (exang):

-   Over a third of patients (36.63%) experience exercise-induced angina, indicating its relevance as a risk factor.

-   The majority (57.39%) do not have this symptom, suggesting varied clinical presentations.

-   This distribution warrants further analysis to link exang with heart disease severity

**5.2 Chest Pain Type (cp)**

-   We will analyze the distribution of chest pain types. The categories are:

    -   **0**: Typical Angina

    -   **1**: Atypical Angina

    -   **2**: Non-Anginal Pain

    -   **3**: Asymptomatic

```{r}
# Count of patients in each chest pain type
cp_counts <- table(heart_data$cp, useNA = "ifany")
cp_percentage <- prop.table(cp_counts) * 100

# Check unique values in cp column
cat("Unique values in cp column:", unique(heart_data$cp), "\n\n")

# Define chest pain labels (adjust based on actual data)
cp_labels <- c(
  "typical angina" = "Typical Angina",
  "atypical angina" = "Atypical Angina",
  "non-anginal" = "Non-Anginal Pain",
  "asymptomatic" = "Asymptomatic"
)

# Display counts and percentages for chest pain types
cat("Chest Pain Type (cp)\n")
for (label in names(cp_counts)) {
  if (label %in% names(cp_labels)) {
    cat(sprintf("%s: %d patients (%.2f%%)\n", 
                cp_labels[label], cp_counts[label], cp_percentage[label]))
  } else {
    cat(sprintf("Unexpected label %s: %d patients (%.2f%%)\n", 
                label, cp_counts[label], cp_percentage[label]))
  }
}
```

```{r}
# Plot for Chest Pain Type (cp)
ggplot(data = heart_data, aes(x = cp, fill = cp)) +
  geom_bar() +
  scale_fill_manual(values = c("typical angina" = "#66C2A5", 
                              "atypical angina" = "#FC8D62", 
                              "non-anginal" = "#8DA0CB", 
                              "asymptomatic" = "#E78AC3")) +
  labs(
    title = "Distribution of Chest Pain Types",
    x = "Chest Pain Type (cp)",
    y = "Number of Patients"
  ) +
  scale_x_discrete(labels = cp_labels) +
  theme_minimal() +
  theme(legend.position = "none")
```

Chest Pain Type (cp):

-   Asymptomatic chest pain dominates (53.91%), potentially complicating diagnosis without other symptoms.

-   Non-anginal and atypical angina are less common, highlighting diverse pain profiles.

-   Standardizing cp labels is critical to ensure reliable modeling and interpretation.

**5. Imputing missing Values**

```{r}
# The following code segment is generated Using AI using the prompt:
# provide a complete code of imputing missing values. Use Mice imputation where the missing values is greater than 30% in any of the column. 
# Ensure heart_data is loaded
# heart_data <- read.csv("heart_disease_data.csv") # Uncomment if needed

# Install and load required packages
if (!require(mice)) install.packages("mice")
library(mice)

# Function to calculate mode for categorical/binary variables
get_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}

# Step 1: Check initial missing values
missing_counts <- colSums(is.na(heart_data))
missing_cols <- names(missing_counts[missing_counts > 0])
cat("Initial missing values:\n")
print(missing_counts[missing_cols])

# Step 2: Separate high and low missingness columns (>30% of 920 rows = 276)
threshold <- 0.3 * nrow(heart_data)  # 30% of 920 ≈ 276
high_missing_cols <- names(missing_counts[missing_counts > threshold])
low_missing_cols <- names(missing_counts[missing_counts > 0 & missing_counts <= threshold])
cat("\nHigh missingness (>30% or >", round(threshold), "):", high_missing_cols, "\n")
cat("Low missingness (≤30% or ≤", round(threshold), "):", low_missing_cols, "\n\n")

# Step 3: Impute low missingness columns
continuous_low <- c("trestbps", "chol", "thalch", "oldpeak")
binary_low <- c("fbs", "exang")
categorical_low <- c("restecg")

# Impute continuous variables with median
for (var in continuous_low) {
  if (var %in% names(heart_data) && any(is.na(heart_data[[var]]))) {
    median_val <- median(heart_data[[var]], na.rm = TRUE)
    heart_data[[var]][is.na(heart_data[[var]])] <- median_val
    cat(sprintf("Imputed %s with median: %.2f\n", var, median_val))
  }
}

# Impute binary variables with mode
for (var in binary_low) {
  if (var %in% names(heart_data) && any(is.na(heart_data[[var]]))) {
    mode_val <- get_mode(heart_data[[var]])
    heart_data[[var]][is.na(heart_data[[var]])] <- mode_val
    cat(sprintf("Imputed %s with mode: %s\n", var, mode_val))
  }
}

# Impute categorical variables with mode
for (var in categorical_low) {
  if (var %in% names(heart_data) && any(is.na(heart_data[[var]]))) {
    mode_val <- get_mode(heart_data[[var]])
    heart_data[[var]][is.na(heart_data[[var]])] <- mode_val
    cat(sprintf("Imputed %s with mode: %s\n", var, mode_val))
  }
}

# Step 4: Ensure slope and thal are factors for MICE
if ("slope" %in% names(heart_data)) {
  heart_data$slope <- as.factor(heart_data$slope)
}
if ("thal" %in% names(heart_data)) {
  heart_data$thal <- as.factor(heart_data$thal)
}

# Step 5: Impute high missingness columns with MICE
if (length(high_missing_cols) > 0) {
  # Initialize MICE to get predictor matrix and methods
  init <- mice(heart_data, maxit = 0)
  meth <- init$method
  pred_matrix <- init$predictorMatrix
  
  # Specify imputation methods for high-missingness columns based on their type
  for (col in high_missing_cols) {
    if (is.factor(heart_data[[col]]) || is.character(heart_data[[col]])) {
      meth[col] <- "polyreg"  # Use polytomous regression for categorical
    } else if (is.numeric(heart_data[[col]])) {
      meth[col] <- "pmm"      # Use predictive mean matching for numeric
    } else {
      meth[col] <- ""         # Skip or add another method if needed
    }
  }
  
  # Ensure other variables are not imputed (only impute high missingness columns)
  meth[setdiff(names(heart_data), high_missing_cols)] <- ""
  
  # Exclude id from predictors (non-predictive) if it exists
  if ("id" %in% names(heart_data)) {
    pred_matrix[, "id"] <- 0
  }
  
  # Run MICE with verbose output for debugging
  imputed_high <- mice(
    heart_data,
    m = 5,
    method = meth,
    predictorMatrix = pred_matrix,
    seed = 123,
    maxit = 10,
    printFlag = TRUE
  )
  
  # Replace high missingness columns with imputed values
  heart_data[, high_missing_cols] <- complete(imputed_high)[, high_missing_cols]
  cat("\nImputed high missingness columns with MICE:\n", high_missing_cols, "\n")
}

# Step 6: Verify imputation
cat("\nMissing values after imputation:\n")
print(colSums(is.na(heart_data)))

# Save the imputed dataset (optional)
# write.csv(heart_data, "heart_data_imputed.csv", row.names = FALSE)

```

```{r}
# Remove the patient Id as it is not need for analysis
heart_data <- heart_data[, !names(heart_data) %in% "id"]
```

**6. Target variable Transformation**

As we saw in previous section that , our target variable `num` is categorical variable with multiple categories like 0, 1, 2, 3, 4 where 0 shows no disease and 1-4 show diseases with different categories. Now We want to make it binary, so we will assign a single category like 1 if `num` is either 1, 2, 3 or 4.

```{r}
# Original num: 0 = no disease, 1-4 = disease severity
heart_data$num_binary <- ifelse(heart_data$num %in% c(1, 2, 3, 4), 1, 0)

# Verify the transformation
cat("Original num distribution:\n")
print(table(heart_data$num))
cat("\nNew binary num distribution:\n")
print(table(heart_data$num_binary))
```

```{r}
# Visualize the binary distribution
ggplot(data = heart_data, aes(x = as.factor(num_binary), fill = as.factor(num_binary))) +
  geom_bar() +
  scale_fill_manual(values = c("0" = "#66C2A5", "1" = "#FC8D62"), 
                    labels = c("No Disease", "Disease")) +
  labs(
    title = "Distribution of Binary Heart Disease Status",
    x = "Heart Disease Status",
    y = "Number of Patients"
  ) +
  scale_x_discrete(labels = c("0" = "No Disease", "1" = "Disease")) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
# Remove the patient Id as it is not need for analysis
heart_data <- heart_data[, !names(heart_data) %in% "num"]
```

So after making it binary, the distribution now seems fine. So we don't need to oversample it for class imbalance.

**7. Encoding Categorical variables**

```{r}
dummy <- dummyVars(" ~ sex + dataset + fbs + exang + cp + restecg + slope + thal + ca", data = heart_data)
dummy_df <- predict(dummy, heart_data)
heart_data <- cbind(heart_data[, !names(heart_data) %in% c("cp", "restecg", "slope", "thal", "ca")], dummy_df)
```

```{r}
dim(heart_data)
```

**8. Data Splitting**

split the data into 80% training and 20% testing

```{r}
inTrain <- createDataPartition(heart_data$num, p=0.8, list=FALSE)
heart_train <- heart_data[inTrain, ]
heart_test <- heart_data[-inTrain, ]
```

```{r}
dim(heart_train)
```

```{r}
dim(heart_test)
```

Further split the training data into 90% training and 10% validation

```{r}
set.seed(1)
inTrainVal <- createDataPartition(heart_train$num, p=0.9, list=FALSE)
heart_train_final <- heart_train[inTrainVal, ]
heart_val <- heart_train[-inTrainVal, ]
```

**9. Create a benchmark model**

```{r}
# Create the Benchmark Classifier

# Compute the majority class in the training data (heart_train_final)
majority_class <- names(which.max(table(heart_train_final$num_binary)))
cat("Majority class in the training data:", majority_class, "\n")

# Create predictions that always predict the majority class
benchmark_preds <- rep(majority_class, nrow(heart_train_final))

# For evaluation, force the positive class to be "1" (assuming 1 = disease)
benchmark_conf_matrix <- confusionMatrix(
  as.factor(benchmark_preds), 
  as.factor(heart_train_final$num_binary),
  positive = "1"
)

cat("Benchmark Classifier Performance (Majority-Class Prediction):\n")
print(benchmark_conf_matrix)

# Compute Evaluation Metrics
# Extract Accuracy from the confusion matrix
accuracy <- benchmark_conf_matrix$overall["Accuracy"]

# Extract precision and recall for the positive class:
# - Precision (Pos Pred Value) is the proportion of predicted positives that are true positives.
# - Recall (Sensitivity) is the proportion of actual positives that are correctly predicted.
precision <- benchmark_conf_matrix$byClass["Pos Pred Value"]
recall <- benchmark_conf_matrix$byClass["Sensitivity"]

# Calculate the F1 Score (harmonic mean of precision and recall)
if((precision + recall) == 0) {
  f1 <- 0
} else {
  f1 <- 2 * (precision * recall) / (precision + recall)
}


# Compute ROC-AUC
# For the ROC, we need predicted probabilities.
# Since our benchmark always predicts the majority class,
# assign probability 1 if majority_class is "1" (i.e., always predict disease), else 0.
benchmark_probs <- if (majority_class == "1") rep(1, nrow(heart_train_final)) else rep(0, nrow(heart_train_final))

# Compute ROC curve and AUC.
# Specify the positive level as "1". Also, setting the levels explicitly ensures correct ordering.
roc_obj <- roc(response = heart_train_final$num_binary, 
               predictor = benchmark_probs, 
               levels = c("0", "1"))
auc_benchmark <- auc(roc_obj)


# Print All Metrics

cat("\n--- Benchmark Classifier Metrics ---\n")
cat(sprintf("Accuracy : %.4f\n", accuracy))
cat(sprintf("Precision: %.4f\n", precision))
cat(sprintf("Recall   : %.4f\n", recall))
cat(sprintf("F1 Score : %.4f\n", f1))
cat(sprintf("ROC AUC  : %.4f\n", auc_benchmark))

```

**10. Train ML Model**

**10.1 Lasso Regression**

```{r}
heart_train_final$num_binary <- as.factor(heart_train_final$num_binary)
heart_test$num_binary <- as.factor(heart_test$num_binary)
heart_val$num_binary <- as.factor(heart_val$num_binary)

```

```{r}
set.seed(1)
```

```{r}
# Define 5-fold cross-validation control for models (using Kappa as metric)
ctrl <- trainControl(method = "cv", 
                     number = 5, 
                     verboseIter = TRUE)

# Define training control with data normalization for SVM models (center & scale)
train_control_norm <- trainControl(
  method = "cv",
  number = 5,
  verboseIter = TRUE,
  preProc = c("center", "scale")
)
```

```{r}
grid_lasso <- expand.grid(
  alpha = 1,          # Lasso regression (elastic net mixing parameter)
  lambda = seq(0.01, 1000, length = 100)  # Range of lambda values from 0.01 to 1000
)
```

```{r}
lasso_model <- train(num_binary ~ ., 
                     data = heart_train_final, 
                     method = "glmnet",
                     trControl = ctrl,
                     tuneGrid = grid_lasso,
                     metric = "Kappa",
                     family = "binomial"
)

```

```{r}
lasso_model
```

```{r}
# Check the best lambda
best_lambda_lasso <- lasso_model$bestTune$lambda
cat("Best lambda for LASSO:", best_lambda_lasso, "\n")
```

**10.2 Ridge Regression**

```{r}
# Define tuning grid for Ridge (alpha = 0)
grid_ridge <- expand.grid(alpha = 0, lambda = 10^seq(3, -2, length = 100))

ridge_model <- train(num_binary ~ ., 
                     data = heart_train_final, 
                     method = "glmnet",
                     trControl = ctrl,
                     tuneGrid = grid_ridge,
                     metric = "Kappa",
                     family = "binomial"
)

```

```{r}
ridge_model
```

```{r}
# Check the best lambda for Ridge
best_lambda_ridge <- ridge_model$bestTune$lambda
cat("Best lambda for Ridge:", best_lambda_ridge, "\n")
```

```{r}
# Get coefficients for the best tuned Ridge model
coefficients_ridge <- coef(ridge_model$finalModel, s = best_lambda_ridge)
print(coefficients_ridge)
```

**10.3 Elastic Net Regression**

```{r}
# Define tuning grid for Elastic Net (alpha ranges from 0 to 1)
grid_en <- expand.grid(alpha = seq(0, 1, by = 0.1), lambda = 10^seq(3, -2, length = 100))

elastic_net_model <- train(num_binary ~ ., 
                           data = heart_train_final, 
                           method = "glmnet",
                           trControl = ctrl,
                           tuneGrid = grid_en,
                           metric = "Kappa",
                           family = "binomial"
)
```

```{r}
elastic_net_model
```

```{r}
# Get best tuning parameters for Elastic Net
best_alpha_en <- elastic_net_model$bestTune$alpha
best_lambda_en <- elastic_net_model$bestTune$lambda
cat("Best Alpha for Elastic Net:", best_alpha_en, "\n")
cat("Best Lambda for Elastic Net:", best_lambda_en, "\n")

```

```{r}
# Get coefficients for the best tuned Elastic Net model
coefficients_en <- coef(elastic_net_model$finalModel, s = best_lambda_en)
print(coefficients_en)
```

**10.4 Random Forest model**

```{r}
rf_model <- train(num_binary ~ ., 
                  data = heart_train_final, 
                  method = "rf",
                  trControl = ctrl,
                  importance = TRUE,   # Compute variable importance
                  metric = "Kappa"
)
```

```{r}
rf_model
```

```{r}
# Get and plot variable importance
var_importance <- varImp(rf_model)
print(var_importance)
plot(var_importance)
```

**10.5 Gradient Boosting Model**

```{r}
gbm_model <- train(num_binary ~ ., 
                   data = heart_train_final, 
                   method = "gbm",
                   trControl = ctrl,
                   metric = "Kappa",
                   verbose = FALSE
)
```

```{r}
gbm_model
```

**10.6 SVM Model**

**10.6.1 SVM with Linear Kernel**

```{r}
svm_linear_model <- train(num_binary ~ ., 
                          data = heart_train_final, 
                          method = "svmLinear",
                          trControl = train_control_norm,
                          metric = "Kappa"
)

```

```{r}
svm_linear_model
```

**10.6.2 SVM with Radial Kernel**

```{r}
svm_radial_model <- train(num_binary ~ ., 
                          data = heart_train_final, 
                          method = "svmRadial",
                          trControl = train_control_norm,
                          metric = "Kappa"
)

```

```{r}
svm_radial_model
```

**11. Summary of all the models**

```{r}
# Combine models into a list for comparison
models_list <- list(
  LASSO = lasso_model,
  RIDGE = ridge_model,
  Elastic_Net = elastic_net_model,
  Random_Forest = rf_model,
  GBM = gbm_model,
  SVM_Linear = svm_linear_model,
  SVM_Radial = svm_radial_model
)

# Compare models using resamples
model_resamples <- resamples(models_list)
summary(model_resamples)
```

**12 Model Evaluations on test set**

```{r}
evaluate_model <- function(model, data, true_labels, model_name = "Model") {
  # Generate predicted class labels
  preds <- predict(model, newdata = data)
  
  # Generate predicted probabilities for class "1" (if supported)
  if ("prob" %in% model$modelType || "prob" %in% names(model)) {
    prob_preds <- predict(model, newdata = data, type = "prob")[, "1"]
  } else {
    prob_preds <- as.numeric(as.character(preds))
  }
  
  # Confusion Matrix
  cm <- confusionMatrix(preds, true_labels, positive = "1")
  
  # Extract Metrics
  acc <- cm$overall["Accuracy"]
  prec <- cm$byClass["Pos Pred Value"]
  rec <- cm$byClass["Sensitivity"]
  f1 <- if ((prec + rec) == 0) 0 else 2 * (prec * rec) / (prec + rec)
  
  # ROC AUC
  roc_obj <- roc(response = true_labels, predictor = prob_preds, levels = c("0", "1"))
  auc_val <- auc(roc_obj)
  
  # Print all
  cat(sprintf("\n--- %s ---\n", model_name))
  cat(sprintf("Accuracy : %.4f\n", acc))
  cat(sprintf("Precision: %.4f\n", prec))
  cat(sprintf("Recall   : %.4f\n", rec))
  cat(sprintf("F1 Score : %.4f\n", f1))
  cat(sprintf("ROC AUC  : %.4f\n", auc_val))
}

```

```{r}
# Set factor levels just in case
heart_test$num_binary <- factor(heart_test$num_binary, levels = c("0", "1"))

# Evaluate Models on the Test Set
evaluate_model(lasso_model, heart_test, heart_test$num_binary, "LASSO Regression")
evaluate_model(ridge_model, heart_test, heart_test$num_binary, "Ridge Regression")
evaluate_model(elastic_net_model, heart_test, heart_test$num_binary, "Elastic Net Regression")
evaluate_model(rf_model, heart_test, heart_test$num_binary, "Random Forest")
evaluate_model(gbm_model, heart_test, heart_test$num_binary, "Gradient Boosting")
evaluate_model(svm_linear_model, heart_test, heart_test$num_binary, "SVM Linear")
evaluate_model(svm_radial_model, heart_test, heart_test$num_binary, "SVM Radial")

```

**13. Neural Network Model**

**13.1 Scale Numerical Columns**

```{r}
num_vars <- c("age", "trestbps", "chol", "thalch", "oldpeak")
train_means <- colMeans(heart_train_final[, num_vars])
train_sds <- apply(heart_train_final[, num_vars], 2, sd)

heart_train_final[, num_vars] <- scale(heart_train_final[, num_vars], center=train_means, scale=train_sds)
heart_val[, num_vars] <- scale(heart_val[, num_vars], center=train_means, scale=train_sds)
heart_test[, num_vars] <- scale(heart_test[, num_vars], center=train_means, scale=train_sds)
```

**13.2 Prepare data for NN Model**

```{r}
X_train <- subset(heart_train_final, select = -num_binary)
y_train <- heart_train_final$num_binary

X_val <- subset(heart_val, select = -num_binary)
y_val <- heart_val$num_binary

X_test <- subset(heart_test, select = -num_binary)
y_test <- heart_test$num_binary
```

```{r}
# Convert feature sets to numeric matrices using model.matrix (remove intercept)
X_train_mat <- model.matrix(~ . - 1, data = X_train)
X_val_mat   <- model.matrix(~ . - 1, data = X_val)
X_test_mat  <- model.matrix(~ . - 1, data = X_test)

# Confirm that the matrices are numeric
str(X_train_mat)

# If your target variables are factors (e.g., "0", "1"), convert them to numeric
y_train_numeric <- as.numeric(as.character(y_train))
y_val_numeric   <- as.numeric(as.character(y_val))
y_test_numeric  <- as.numeric(as.character(y_test))
```

```{r}
cat("X_train dimensions:", dim(X_train), "\n")
cat("y_train length:", length(y_train), "\n\n")

cat("X_val dimensions:", dim(X_val), "\n")
cat("y_val length:", length(y_val), "\n\n")

cat("X_test dimensions:", dim(X_test), "\n")
cat("y_test length:", length(y_test), "\n")
```

**13.3 Simple Neural Network Model**

```{r}
model <- keras_model_sequential() %>% 
  # Hidden layer 1: Input shape is the number of columns in X_train_mat
  layer_dense(units = 128, activation = "relu", input_shape = dim(X_train_mat)[2]) %>%
  layer_dropout(rate = 0.3) %>%
  
  # Hidden layer 2
  layer_dense(units = 64, activation = "relu") %>%
  layer_dropout(rate = 0.4) %>% 
  
  # Hidden layer 3
  layer_dense(units = 64, activation = "relu") %>% 
  layer_dropout(rate = 0.4) %>% 
  
  # Output layer: one unit for binary classification using sigmoid activation
  layer_dense(units = 1, activation = "sigmoid")

# Compile the model with binary cross-entropy loss and Adam optimizer
model %>% compile(
  loss = "binary_crossentropy", 
  optimizer = optimizer_adam(),
  metrics = c("accuracy")
)

# Display model summary
summary(model)
```

```{r}
history <- model %>% fit(
  X_train_mat, 
  y_train_numeric, 
  batch_size = 64, 
  epochs = 20, 
  validation_data = list(X_val_mat, y_val_numeric), 
  callbacks = list(callback_early_stopping(monitor = "loss", patience = 10, restore_best_weights = TRUE))
)
```

```{r}
plot(history)
```

**13.4 Tuning the hyper-parameters**

```{r}
# Define hyperparameter grid for tuning
params <- list(
  learning_rate = c(0.0001, 0.001, 0.01),
  dropout1 = c(0.3, 0.4, 0.5),
  dropout2 = c(0.3, 0.4, 0.5),
  dropout3 = c(0.3, 0.4, 0.5),
  neurons1 = c(64, 128, 256),
  neurons2 = c(64, 128, 256),
  neurons3 = c(64, 128, 256),
  l2 = c(0.0001, 0.001, 0.01)
)

# Run tuning (adjust sample % as needed)
set.seed(1)
runs <- tuning_run("tuning_script_nn.R", flags = params, sample = 0.001)
```

```{r}
print(runs)
```

```{r}
view_run(runs$run_dir[3])  
```

Following are the best hyperparameters after tuning:

|  **Hyperparameter**                |  **Value**                         |
|------------------------------------|------------------------------------|
|  learning_rate                     |  0.01                              |
|  dropout1                          |  0.3                               |
|  dropout2                          |  0.5                               |
|  dropout3                          |  0.5                               |
|  neurons1                          |  64                                |
|  neurons2                          |  64                                |
|  neurons3                          |  256                               |
|  l2                                |  0.0001                            |
|  optimizer                         |  RMSprop                           |
|  epochs                            |  10                                |

**13.5 Retrain the best model on combined data**

```{r}
# Train final model on combined training + validation data
X_combined <- rbind(X_train_mat, X_val_mat)
y_combined <- c(y_train_numeric, y_val_numeric)

```

```{r}
final_model <- keras_model_sequential() %>% 
  layer_dense(
    units = 64, 
    activation = "relu", 
    input_shape = dim(X_combined)[2],
    kernel_regularizer = regularizer_l2(l = 0.0001)
  ) %>%
  layer_dropout(rate = 0.3) %>%
  
  layer_dense(
    units = 64, 
    activation = "relu"
  ) %>%
  layer_dropout(rate = 0.5) %>%
  
  layer_dense(
    units = 256, 
    activation = "relu"
  ) %>%
  layer_dropout(rate = 0.5) %>%
  
  layer_dense(units = 1, activation = "sigmoid")

# Compile the model with the tuned learning rate and optimizer
final_model %>% compile(
  loss = "binary_crossentropy", 
  optimizer = optimizer_rmsprop(learning_rate = 0.01),
  metrics = c("accuracy")
)

# Train the model using the combined data
history_final <- final_model %>% fit(
  X_combined, 
  y_combined, 
  batch_size = 64, 
  epochs = 10,  # Based on tuning result
  callbacks = list(callback_early_stopping(monitor = "loss", patience = 10, restore_best_weights = TRUE))
)

```

**13.6 Print all the metrics for NN Model**

```{r}
# Predict on test set
pred_probs <- final_model %>% predict(X_test_mat)
predictions <- ifelse(pred_probs > 0.5, 1, 0)

# Confusion matrix and metrics
conf_matrix <- confusionMatrix(
  factor(predictions, levels = c(0,1)),
  factor(y_test_numeric, levels = c(0,1))
)

accuracy <- conf_matrix$overall['Accuracy']
precision <- conf_matrix$byClass['Precision']
recall <- conf_matrix$byClass['Recall']
f1_score <- conf_matrix$byClass['F1']

# Print metrics
print(conf_matrix)
cat("\nAccuracy:", accuracy)
cat("\nPrecision:", precision)
cat("\nRecall:", recall)
cat("\nF1 Score:", f1_score)

# ROC and AUC
roc_obj <- roc(as.numeric(as.character(y_test)), as.numeric(pred_probs))
auc_value <- auc(roc_obj)
cat("\nAUC:", auc_value, "\n")

```

```{r}
plot(roc_obj, col = "darkred", lwd = 2, main = "Neural Network ROC Curve")
```
